---
title: "xcms-based preprocessing of LC-MS/MS data for feature-based molecular networking with GNPS2"
author: "Philippine Louail, Steffen Neumann, Johannes Rainer"
format:
  html:
    minimal: true
knitr:
  opts_chunk:
    message: false
vignette: >
  %\VignetteIndexEntry{xcms-based preprocessing for FBMN with GNPS2}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---


# Introduction

This document describes data inspection and preprocessing of an LC-MS/MS data
set using *xcms* [@louail_xcms_2025] and export of the data for subsequent
feature-based molecular networking (FBMN) with GNPS2. Functionality from
different packages from the RforMassSpectrometry package ecosystem are combined
to visualize and process the data.

For details and more in depth description of the various visualizations and
analysis options as well as parameter choices see also the
[Metabonaut](https://rformassspectrometry.github.io/Metabonaut) tutorials
[@louail_rformassspectrometrymetabonaut_2025].

This analysis and the used settings should be considered *initial* with
potential refinement and improvement based on discussions expected during
integration of the analysis into the FBMN workflow.

# Required software packages

The various software packages required for the analysis are defined and loaded
below. All packages are available through Bioconductor or CRAN and can be
installed with `BiocManager::install(<package name>)`.

```{r, message = FALSE}
library(MsExperiment) # container for MS data
library(Spectra)      # main MS infrastructure for R
library(xcms)         # for preprocessing of LC-MS and LC-MS/MS data
library(MsBackendMgf) # to export MS data in MGF format

library(RColorBrewer) # to define colors
library(pander)       # to format tables
library(pheatmap)     # visualization of clustering results as heatmap
library(vioplot)      # to create *violin plots*
```

# Data import

The data analyzed here is part of the MassIVE
[MSV000090156](https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=06bd49807caa4390961fb827606a8696)
data set. The full data set (raw data files) should first downloaded from the
respective directory of the ftp server:
[ftp://massive-ftp.ucsd.edu/v04/MSV000090156/](ftp://massive-ftp.ucsd.edu/v04/MSV000090156/).
Here we will analyze the data from *Lab 2*. Before loading the data we define a
`data.frame` with sample and experiment-specific information for the individual
MS runs/data files. These should ideally comprise all relevant phenotypic but
also technical information (e.g. injection index) to allow proper adjusting or
modeling of the data.

```{r}
pd <- data.frame(
    file_name = c("Interlab-LC-MS_Lab2_A15M_Pos_MS2_Rep1.mzML",
                  "Interlab-LC-MS_Lab2_A15M_Pos_MS2_Rep2.mzML",
                  "Interlab-LC-MS_Lab2_A15M_Pos_MS2_Rep3.mzML",
                  "Interlab-LC-MS_Lab2_A45M_Pos_MS2_Rep1.mzML",
                  "Interlab-LC-MS_Lab2_A45M_Pos_MS2_Rep2.mzML",
                  "Interlab-LC-MS_Lab2_A45M_Pos_MS2_Rep3.mzML",
                  "Interlab-LC-MS_Lab2_A5M_Pos_MS2_Rep1.mzML",
                  "Interlab-LC-MS_Lab2_A5M_Pos_MS2_Rep2.mzML",
                  "Interlab-LC-MS_Lab2_A5M_Pos_MS2_Rep3.mzML",
                  "Interlab-LC-MS_Lab2_M_Pos_MS2_Rep1.mzML",
                  "Interlab-LC-MS_Lab2_M_Pos_MS2_Rep2.mzML",
                  "Interlab-LC-MS_Lab2_M_Pos_MS2_Rep3.mzML",
                  "Interlab-LC-MS_Lab2_PPL_Pos_MS2_Rep1.mzML"),
    sample_name = c("A15M", "A15M", "A15M",
                    "A45M", "A45M", "A45M",
                    "A5M", "A5M", "A5M",
                    "M", "M", "M",
                    "PPL"),
    sample_desc = c("A15M_R1", "A15M_R2", "A15M_R3",
                    "A45M_R1", "A45M_R2", "A45M_R3",
                    "A5M_R1", "A5M_R2", "A5M_R3",
                    "M_R1", "M_R2", "M_R3",
                    "PPL_R1"),
    replicate = c(1, 2, 3, 1, 2, 3, 1, 2, 3, 1, 2, 3, 1)
)
```

To run this analysis the MS data files (in mzML format) above need to be
available. In the example workflow, they were downloaded from the MassIVE ftp
server and stored to a local folder */data/massive-ftp.ucsd.edu/v04*. Below we
define the path to the data files and load the data set. This needs to be
adapted if the files were stored to a different folder.

```{r}
path <- file.path("/data", "massive-ftp.ucsd.edu", "v04",
                  "MSV000090156", "peak", "mzml", "POS_MSMS",
                  "Lab_2")
mse <- readMsExperiment(file.path(path, pd$file_name),
                        sampleData = pd)
mse
```

The samples and data files from the present data set are displayed in the table
below. Note that in the R code block below we use the R *pipe* operator `|>` to
avoid nested function calls and improve the readability of the code.

```{r, results = "asis"}
sampleData(mse)[, c("sample_name", "sample_desc", "replicate")] |>
    as.data.frame() |>
    pandoc.table(style = "rmarkdown", split.table = Inf)
```

We in addition also define different colors for the individual samples.

```{r}
#' Define a color for each unique original sample
col <- sampleData(mse)$sample_name |>
                     unique() |>
                     length() |>
                     brewer.pal("Set2")
names(col) <- unique(sampleData(mse)$sample_name)

#' Define a color for each data file/sample
col_sample <- col[sampleData(mse)$sample_name]
```

# Data visualization and general quality assessment

For data inspection and a general data overview we first create base peak
chromatograms (BPC) and total ion chromatograms (TIC) of the data set using the
`chromatograms()` functions specifying either `"max"` (for BPC) or `"sum"` (for
TIC) as the function to aggregate the per-spectrum intensities.

```{r}
#' BPC
bpc <- chromatogram(mse, aggregationFun = "max")

plot(bpc, col = paste0(col_sample, 80), main = "BPC", lwd = 2)
grid()
legend("topright", col = col, legend = names(col), lty = 1, lwd = 2)
```

```{r}
#' TIC
tic <- chromatogram(mse, aggregationFun = "sum")

plot(tic, col = paste0(col_sample, 80), main = "TIC", lwd = 2)
grid()
legend("topright", col = col, legend = names(col), lty = 1, lwd = 2)
```

Based on the BPC and TIC there seems to be little retention time shifts between
the samples. Also, no signal seems to be present before 20 seconds and after 850
seconds. Thus, we below filter the data set to spectra acquired within this
retention time range.

```{r}
#' filter the data set to a retention time range from 20 to 850 seconds
mse <- filterSpectra(mse, filterRt, c(20, 850))
```

BPC and TIC aggregate data along the *m/z* dimension per spectrum (retention
time) to compare the signal measured along the retention time. To compare the
*mass* or ion content of the individual samples/measurement runs we in addition
aggregate data along the retention time, for distinct *m/z* values.

To this end we first bin each spectrum to get discrete and similar *m/z* values
within the data set.

```{r}
#' bin mass peaks into into discrete m/z bins of 0.02 Da.
s_bin <- spectra(mse) |>
    filterMsLevel(1L) |>
    bin(binSize = 0.02)

#' combine all spectra within the same sample into a single spectrum
#' reporting the maximum intensity of all mass peaks with the same m/z bin
bps <- combineSpectra(s_bin, f = s_bin$dataOrigin, intensityFun = max)
#' the same but reporting the sum of intensities per m/z bin
tis <- combineSpectra(s_bin, f = s_bin$dataOrigin, intensityFun = sum)
```

We have thus a single aggregated mass spectrum per sample. These *base peak
spectra* are plotted below.

```{r}
#| fig-cap: "Aggregated MS1 spectrum per sample."
#| fig-width: 9.5
#| fig-height: 9.5
par(mar = c(4.5, 4, 0, 0.5))
plotSpectra(bps, main = "")
```

The *mass content* of the samples seems to be similar, with the exception of the
last file. We can also plot all spectra into the same plot, using a different
color per sample.

```{r}
plotSpectraOverlay(bps, col = paste0(col_sample, 40), lwd = 2)
grid()
legend("topright", col = col, legend = names(col), lty = 1)
```

We can also use these aggregated spectra to calculate spectra similarity between
the individual samples and cluster them.

```{r}
#| fig-cap: "Similarity of *base peak spectra* of all samples in the experiment."
#| fig-width: 9.5
#| fig-height: 9.5
sim <- compareSpectra(bps)
rownames(sim) <- colnames(sim) <- sampleData(mse)$sample_desc
pheatmap(sim)
```

A15M, A5M and M samples cluster together, separately from the A45M samples while
the PPL_R1 sample has a distinct mass peak profile.


# Data preprocessing

Data preprocessing is the first step in the analysis of LC-MS data processing
and analyzing the raw MS data to result in a two-dimensional quantification
table of LC-MS *features* in the various samples of the experiment. This process
consists of 3 main steps: **chromatographic peak detection**, **retention time
alignment** and **correspondence analysis**. An additional **gap-filling** step
can be conducted to reduce the number of missing values integrating raw MS
signal from the expected *m/z* by retention time areas of defined LC-MS
features.

Most methods in *xcms* are parallelized by default. Below we define the parallel
processing setup for the present analysis.

::: {.callout-note collapse="true"}
## Parallel processing details

It is generally advisable to configure the parallel processing setup globally
using `register()`. On Unix machines *multi-core* parallel processing can be
used, that shares memory between the parallel processes. Windows supports only
*socket-based* parallel processing and starts a separate R for each parallel
process. Below we globally define the parallel processing setup depending on the
operating system. For the present analysis we use 4 parallel processes.
:::

```{r}
if (.Platform$OS.type == "unix") {
    register(MulticoreParam(4))
} else {
    register(SnowParam(4))
}
```

## Chromatographic peak detection

The aim of the chromatographic peak detection is to identify and quantify signal
in the raw MS data space representing signal from ions of compounds present in
a sample. Data processing is performed separately for each data file and mass
peak intensities with similar *m/z* are evaluated along retention time axis to
identify chromatographic peaks. We use the *centWave* algorithm for peak
detection. The most important parameter for *centWave* is `peakwidth` which
defines an approximate lower and upper expected width of chromatographic peaks
in retention time dimension. Without any prior information, we need to derive
this information from the data set. We therefore zoom into areas of the BPC that
seem to contain signal from an ion.

```{r}
#| fig-cap: "Definition of an example region for EIC extraction. Upper panel: BPC, dashed vertical lines indicate the selected retention time region. Lower panel: EIC for the *m/z* region with the largest signal in the retention time window."
#| fig-width: 9.5
#| fig-height: 9.5

#' extract BPC
par(mfrow = c(2, 1))
bpc <- chromatogram(mse, aggregationFun = "max")
plot(bpc, col = paste0(col_sample, 80), lwd = 2)
grid()

#' identify a retention time region to extract
rtr_1 <- c(105, 130)
abline(v = rtr_1, lty = 2)

#' identify the m/z with the largest intensity in that region:
#' - restrict to MS1 data
#' - filter the MS data by retention time
#' - extract the MS data as a data.frame
tmp <- spectra(mse) |>
    filterMsLevel(1L) |>
    filterRt(rtr_1) |>
    longForm(columns = c("mz", "intensity"))

#' define a m/z range around the m/z with largest intensity
mzr_1 <- tmp$mz[which.max(tmp$intensity)] + c(-0.005, 0.005)

#' extract an EIC for that RT and m/z region
eic_1 <- chromatogram(mse, rt = rtr_1, mz = mzr_1)
plot(eic_1, col = paste0(col_sample, 80), lwd = 2)
grid()
```

The width of this chromatographic peaks is about 8 seconds. We evaluate a second
signal at the and of the chromatogram.

```{r}
#| fig-cap: "Definition of an example region for EIC extraction. Upper panel: BPC, dashed vertical lines indicate the selected retention time region. Lower panel: EIC for the *m/z* region with the largest signal in the retention time window."
#| fig-width: 9.5
#| fig-height: 9.5

par(mfrow = c(2, 1))
plot(bpc, col = paste0(col_sample, 80), lwd = 2)
grid()

#' identify a region to extract
rtr_2 <- c(720, 770)
abline(v = rtr_2, lty = 2)

#' identify the m/z with the largest intensity in that region:
#' - restrict to MS1 data
#' - filter the MS data by retention time
#' - extract the MS data as a data.frame
tmp <- spectra(mse) |>
    filterMsLevel(1L) |>
    filterRt(rtr_2) |>
    longForm(columns = c("mz", "intensity"))

#' define a m/z range around the m/z with largest intensity
mzr_2 <- tmp$mz[which.max(tmp$intensity)] + c(-0.005, 0.005)

#' extract an EIC for that RT and m/z region
eic_2 <- chromatogram(mse, rt = rtr_2, mz = mzr_2)
plot(eic_2, col = paste0(col_sample, 80), lwd = 2)
grid()
```

The width of the chromatographic peaks for that *m/z* slice seem to be around 15
seconds. Also, there seems to be a considerable shift in retention times between
the samples.

Based on these two example signals we define the `peakdwidth` parameter for the
present data set to be between and 5 and 20 seconds. For a *real* data analysis
it is suggested to evaluate more signals, also eventually of internal standards
or ions of compounds that are expected to be present in the sample.

A second potentially data set-specific parameter of the *centWave* algorithm is
`ppm`. It defines the maximum allowed deviation in *m/z* dimension of mass peaks
in consecutive spectra considered to represent signal from the same ion. To
illustrate this we below subset the full MS data from the first data file to the
*m/z* and retention time range defined above and plot the individual mass peaks.

::: {.callout-note appearance="minimal"}
## Details on the `ppm` *centWave* parameter

The `ppm` parameter defines the expected (or observed) *m/z* deviation of mass
peaks representing signal from the same compound/ion in consecutive
spectra. This scattering of mass peak's *m/z* values can depend on the
centroiding algorithm used or the precision of the MS instrument. In addition,
on TOF instruments, the scattering can depend on the intensity of the signal,
with higher variation observed for low intensity peaks and increasing stability
with higher signal.
:::

```{r}
mse[1L] |>
    filterSpectra(filterMsLevel, 1L) |>
    filterSpectra(filterRt, rt = rtr_1) |>
    filterSpectra(filterMzRange, mz = mzr_1) |>
    plot()
```

The individual mass peaks are shown in the lower panel in the plot above. For
the present ion, the *m/z* values show a very low variance.

We evaluate the signal also for the second *m/z* - retention time window defined
above.

```{r}
mse[1L] |>
    filterSpectra(filterMsLevel, 1L) |>
    filterSpectra(filterRt, rt = rtr_2) |>
    filterSpectra(filterMzRange, mz = mzr_2) |>
    plot()
```

The scattering of *m/z* values looks larger, but is still below 0.001 Da. We
nevertheless use a `ppm = 20` for the present data set - as we do not assume
that ions from different compounds elute at the same time with a difference in
ppm lower than 20.

Below we define the settings for *centWave* and perform the chromatographic peak
detection on the full data set. Note that it can also be helpful to test the
different settings by performing peak detection on extracted ion chromatograms
as described in
[Metabonaut](https://rformassspectrometry.github.io/Metabonaut/). Parameter
`chunkSize` defines the number of data files from which MS data should be loaded
at a time. This parameter thus has an influence on the memory usage of the
analysis.

```{r}
#' configure and perform chromatographic peak detection
cwp <- CentWaveParam(
    ppm = 20,
    peakwidth = c(5, 20),
    snthresh = 8,
    integrate = 2,
    mzdiff = 0.001
)
mse <- findChromPeaks(mse, param = cwp, chunkSize = 4L)
```

In most cases it is also advisable to perform a *peak postprocessing* to remove
artifacts of the *centWave* peak detection (e.g. overlapping or split
peaks). Below we perform such peak refinement which will merge partially or
completely overlapping chromatographic peak, or those that are less than 2
seconds apart from each other, if the intensity below both apexes is lower than
a certain proportion of the apex intensity of the lower intensity peak.

```{r}
#' configure and perform *peak refinement*
mnpp <- MergeNeighboringPeaksParam(
    expandRt = 1,
    expandMz = 0,
    ppm = 0,
    minProp = 0.75)
mse <- refineChromPeaks(mse, mnpp, chunkSize = 4L)
```

We next compare the number of identified peaks per sample as well as their *m/z*
and retention time widths.

```{r}
#' split the detected chrom peaks per sample
pk_list <- split.data.frame(
    chromPeaks(mse, columns = c("mzmin", "mzmax", "rtmin", "rtmax")),
    chromPeaks(mse, columns = "sample")[, "sample"])
#' calculate mz and rt widths
pk_list <- lapply(pk_list, function(z) {
    cbind(z, mz_width = z[, "mzmax"] - z[, "mzmin"],
          mz_width_ppm = (z[, "mzmax"] - z[, "mzmin"]) * 1e6 / z[, "mzmax"],
          rt_width = z[, "rtmax"] - z[, "rtmin"])
})

#' plot the information
par(mfrow = c(3, 1), mar = c(0, 4.3, 1.5, 0.1))
barplot(unlist(lapply(pk_list, nrow)),
     col = col_sample,
     ylab = "peak count", main = "Peak detection summary, mse", xaxt = "n")
grid()
legend("top", horiz = TRUE, col = col, pch = 15,
       legend = names(col))
par(mar = c(0, 4.3, 0, 0.1))
vioplot(lapply(pk_list, function(z) z[, "mz_width_ppm"]), outline = FALSE,
        ylab = "m/z width [ppm]", xaxt = "n", line = 3,
        col = col_sample)
grid()
vioplot(lapply(pk_list, function(z) z[, "rt_width"]),
        ylab = "rt width [s]", col = col_sample, line = 3)
grid()
```

The highest number of peaks was detected in the *PPL* sample. Apart from that
sample, the numbers of detected peaks is comparable in the data set. Also the
*m/z* width and the retention time widths. As expected, most of the identified
chromatographic peaks are about 10 seconds wide. Also, their *m/z* width is
below 10 ppm for most peaks, but some show also a larger *m/z* widths.

We also evaluate the peak detection results on the two example *m/z* - retention
time regions. Identified chromatographic peaks will be colored according to the
sample group.

```{r}
#| fig-cap: "Chromatographic peak detection results on the first example EIC."
eic_1 <- chromatogram(mse, mz = mzr_1, rt = rtr_1)

#' define a color for each chromatographic peak
col_peak <- col_sample[chromPeaks(eic_1)[, "sample"]]
plot(eic_1, col = paste0(col_sample, 80),
     peakBg = paste0(col_peak, 10),
     peakCol = paste0(col_peak, 80))
grid()
legend("topright", col = col, lty = 1,
       legend = names(col))
```


```{r}
#| fig-cap: "Chromatographic peak detection results on the second example EIC."
eic_2 <- chromatogram(mse, mz = mzr_2, rt = rtr_2)

#' define a color for each chromatographic peak
col_peak <- col_sample[chromPeaks(eic_2)[, "sample"]]
plot(eic_2, col = paste0(col_sample, 80),
     peakBg = paste0(col_peak, 10),
     peakCol = paste0(col_peak, 80))
grid()
legend("topright", col = col, lty = 1,
       legend = names(col))
```


## Retention time alignment

The aim of the retention time alignment step is to reduce differences observed
in the elution time of compounds between different LC-MS runs. A variety of
methods for this have been proposed and some are also implemented in *xcms*. We
use the most straight forward approach that aligns chromatographic runs based on
retention times of *anchor peaks*, i.e., compounds present in most of the
samples of an experiment. To define these anchor peaks we must however perform
an initial correspondence analysis and group chromatographic peaks with similar
*m/z* and retention time across samples.

We use the *peak density* correspondence method which groups chromatographic
peaks into an LC-MS feature, if their *m/z* difference is smaller than `binSize`
(+ `ppm` of the *m/z*), the retention time of their apex is within one peak of
the *peak density* curve (which smoothness can be configured with parameter
`bw`) and a chromatographic peak is present in at least `minFraction` of at
least one of the sample groups defined with `sampleGroups`. For the initial
correspondence we apply more relaxed settings and consider all samples to be in
the same sample group (since we want to define anchor peaks that are present in
most samples).

::: {.callout-note appearance="simple" collapse="true"}
## Details on `PeakDensityParam` settings

The *peak density* correspondence method can be configured using the
`PeakDensityParam`. For an initial correspondence used for retention time
alignment more relaxed settings can be used, also putting all samples into the
same *sample group*. The important parameters are `bw` and `binSize`. The former
defines the tolerance of retention time similarity, the latter the similarity of
*m/z* values for chromatographic peaks from different samples to be considered
to represent signal from ions of the same compound. Setting `binSize` is
straight forward - it depends on the resolution of the instrument and the
expected similarity of *m/z* values for the same ion. We set it to a value of
`binSize = 0.01`, hence all chromatographic peaks with a difference in *m/z*
smaller than 0.01 would be evaluated. The `bw` parameter is a bit more difficult
to define. It should be defined based on observed data in the experiment,
ideally, on EICs of closely eluting compounds with the same *m/z*. In our
example we use the second example *m/z* - retention time range, because it
seemed to contain signal from the same compound, but with quite large shifts
between data files. We below *simulate* a correspondence analysis on this EIC.
Other parameters defined are `minFraction`, which defines the minimum required
proportion of samples of one of the sample groups defined with parameter
`sampleGroups` in which a chromatographic peak is present, and `ppm` which,
together with `binSize` defines the maximal accepted difference in
chromatographic peaks' *m/z* values to consider them for grouping.
:::

```{r}
#| fig-cap: "Correspondence analysis simulation on the second example EIC."
pdp <- PeakDensityParam(
    sampleGroups = rep(1, length(mse)),
    bw = 2,
    minFraction = 0.2,
    binSize = 0.01,
    ppm = 10)

col_peak <- col_sample[chromPeaks(eic_2)[, "sample"]]
plotChromPeakDensity(eic_2, param = pdp, col = col_sample, peakCol = col_peak,
                     peakBg = paste0(col_peak, 40))
grid()
```

The upper panel of this plot shows the EIC, the lower panel the data considered
for correspondence: it shows the retention time of the apex positions of all
chromatographic peaks in that *m/z* slice on the x-axis against the sample in
which the chromatographic peak was detected on the y-axis. The black solid line
represents the *peak density estimate*, calculated based on the retention times
of the chromatographic peaks and parameter `bw` with larger values of `bw`
resulting in more smooth curves.

With the used settings, in particular parameter `bw`, the present
chromatographic peaks would be split into two separate LC-MS features (indicated
with a grey rectangle in the lower panel). Assuming that all peaks in that
region represent signal from ions of the same compound, we however want to group
them into the same feature. We hence next increase the `bw` parameter and
simulate the correspondence with these updated settings.

```{r}
#| fig-cap: "Correspondence analysis simulation with `bw = 5`."
pdp@bw <- 5
plotChromPeakDensity(eic_2, param = pdp, col = col_sample, peakCol = col_peak,
                     peakBg = paste0(col_peak, 40))
grid()
```

Changing `bw` to 5 changed the density curve, but we still defined two separate
features. We thus increase below `bw` to 7.

```{r}
#| fig-cap: "Correspondence analysis simulation with `bw = 7`."
pdp@bw <- 7
plotChromPeakDensity(eic_2, param = pdp, col = col_sample, peakCol = col_peak,
                     peakBg = paste0(col_peak, 40))
grid()
```

With a `bw = 7` a single feature was defined. We use these parameter for the
initial correspondence analysis on the full data set.

```{r}
#' perform initial correspondence analysis to group chromatographic peaks
mse <- groupChromPeaks(mse, param = pdp)
```

For the retention time alignment we use the before mentioned *peak groups*
method that aligns LC runs by minimizing retention time differences of so called
anchor peaks. This alignment method is very robust and also flexible, allowing
for example to align samples based on within-experiment QC samples, or against
external reference data or based on manually defined anchor peaks. See
[Metabonaut](https://rformassspectrometry.github.io/Metabonaut) for examples and
options. The method can be configured with `PeakGroupsParam()`. With
`minFraction = 0.9` we below define anchor peaks as those LC-MS features
(defined by the initial correspondence analysis) for which a chromatographic
peak was identified in 90% of the samples of the whole experiment. The observed
retention time differences of these are used to model a curve along retention
time dimension which is then used to align the retention times of all
samples. The smoothness of this curve can be configured with parameter `span`
(values between 0 and 1; values around 0.5 work in most cases).

```{r}
#' configure and run retention time alignment
pgp <- PeakGroupsParam(
    minFraction = 0.90,
    span = 0.4)
mse <- adjustRtime(mse, param = pgp)
```

The effect of this alignment can be visualized with the `plotAdjustedRtime()`
function. It plots the adjusted retention times of each sample on the x-axis
against the difference between the adjusted and raw retention times on the
y-axis as a solid line. Retention times of anchor peaks in each sample are
indicated with individual data points. These should ideally be placed along the
full retention time range of the experiment.

```{r}
#| fig-cap: "Retention time alignment results."
#' visualize alignment results
plotAdjustedRtime(mse, col = paste0(col_sample, 80),
                  peakGroupsPch = 21, lwd = 2)
grid()
legend("topleft", col = col, lty = 1,
       legend = names(col))
```

Anchor points span the full retention time range. Retention time adjustments for
most samples were below 2-4 seconds, with the exception of 3 samples for which a
considerably larger adjustment is present after 500 seconds, and the PPL sample
with larger retention time differences after 650 seconds.

We next evaluate the alignment results based on the BPC before and after
alignment.

```{r}
#| fig-cap: "BPC before (top) and after (bottom) retention time alignment."
#| fig-width: 9.5
#| fig-height: 9.5
#' create a BPC after adjustment; chromPeaks = "none" only creates the BPC
#' without extracting also identified chromatographic peaks.
bpc_adj <- chromatogram(mse, chromPeaks = "none", aggregationFun = "max")

par(mfrow = c(2, 1))
plot(bpc, col = paste0(col_sample, 80), main = "BPC, raw", lwd = 2)
grid()
legend("topright", col = col, legend = names(col), lty = 1, lwd = 2)
plot(bpc_adj, col = paste0(col_sample, 80), main = "BPC, adjusted", lwd = 2)
grid()
```

Misalignment of signal at the later stage of the chromatography seems to be
reduced. We in addition evaluate the effect of retention time alignment on the
two example EICs.

```{r}
#| fig-cap: "First example EIC before (top) and after (bottom) retention time alignment."
#| fig-width: 9.5
#| fig-height: 9.5
eic_1_adj <- chromatogram(mse, rt = rtr_1, mz = mzr_1, chromPeaks = "none")

par(mfrow = c(2, 1))
plot(eic_1, col = paste0(col_sample, 80), lwd = 2, peakType = "none")
grid()
legend("topright", col = col, legend = names(col), lty = 1, lwd = 2)
plot(eic_1_adj, col = paste0(col_sample, 80), lwd = 2)
grid()
```

For this early retention time range already the raw signal was well aligned.

```{r}
#| fig-cap: "Second example EIC before (top) and after (bottom) retention time alignment."
#| fig-width: 9.5
#| fig-height: 9.5
eic_2_adj <- chromatogram(mse, rt = rtr_2, mz = mzr_2, chromPeaks = "none")

par(mfrow = c(2, 1))
plot(eic_2, col = paste0(col_sample, 80), lwd = 2, peakType = "none")
grid()
legend("topright", col = col, legend = names(col), lty = 1, lwd = 2)
plot(eic_2_adj, col = paste0(col_sample, 80), lwd = 2)
grid()
```

This later retention time range shows clear, and strong, differences in
retention times of 4 samples. While the *PPL* sample could be aligned quite well
using the above settings, 3 samples still show considerable shifts in retention
times. We thus re-perform the alignment reducing the value for the `span`
parameter to switch to a more *local* alignment of the samples. We below first
*undo* the retention time alignment, re-perform the initial correspondence
analysis and perform the alignment with the changed settings for parameter
`span`.

```{r}
#' remove retention time alignment results
mse <- dropAdjustedRtime(mse)
#' re-perform initial correspondence
mse <- groupChromPeaks(mse, param = pdp)
#' perform the alignment using updated settings
pgp <- PeakGroupsParam(
    minFraction = 0.90,
    span = 0.1)
mse <- adjustRtime(mse, param = pgp)
```

Evaluating the impact of changing this parameter.

```{r}
#| fig-cap: "Retention time alignment results."
#' visualize alignment results
plotAdjustedRtime(mse, col = paste0(col_sample, 80),
                  peakGroupsPch = 21, lwd = 2)
grid()
legend("topleft", col = col, lty = 1,
       legend = names(col))
```

A stronger alignment can be observed for the retention time area from 750 to 800
seconds. The results for the first example EIC did not change.

```{r}
#| fig-cap: "First example EIC before (top) and after (bottom) retention time alignment."
#| fig-width: 9.5
#| fig-height: 9.5
eic_1_adj <- chromatogram(mse, rt = rtr_1, mz = mzr_1)

par(mfrow = c(2, 1))
#' Setting peakType = "none" prevents identified chromatographic peaks to be
#' indicated in the plot.
plot(eic_1, col = paste0(col_sample, 80), lwd = 2, peakType = "none")
grid()
legend("topright", col = col, legend = names(col), lty = 1, lwd = 2)
plot(eic_1_adj, col = paste0(col_sample, 80), lwd = 2, peakType = "none")
grid()
```

While for the second EIC the alignment improved.

```{r}
#| fig-cap: "Second example EIC before (top) and after (bottom) retention time alignment."
#| fig-width: 9.5
#| fig-height: 9.5
eic_2_adj <- chromatogram(mse, rt = rtr_2, mz = mzr_2)

par(mfrow = c(2, 1))
plot(eic_2, col = paste0(col_sample, 80), lwd = 2, peakType = "none")
grid()
legend("topright", col = col, legend = names(col), lty = 1, lwd = 2)
plot(eic_2_adj, col = paste0(col_sample, 80), lwd = 2, peakType = "none")
grid()
```

Note that in most cases it is not necessary that all samples are perfectly
aligned. Some variation in retention time can be accounted for in the final
correspondence analysis.

## Correspondence analysis

The correspondence analysis groups chromatographic peaks from different samples,
all assumed to represent signal from ions of the same compound, into LC-MS
features. Signals are generally grouped together based on similarity of their
*m/z* and retention times. We use the *peak density* method and, similarly to
what we did in the previous section, we first evaluate the impact of different
parameter settings on the expected results. Also, for the final correspondence
we use the samples' *sample_name* for parameter `sampleGroup`. Combined with
setting `minFraction = 0.67`, an LC-MS feature is defined if a chromatographic
peak is present in at least 67% of the replicates per sample (i.e., 2 out of 3).

```{r}
#' configure the *peak density* correspondence method
pdp <- PeakDensityParam(
    sampleGroups = sampleData(mse)$sample_name,
    minFraction = 0.67,
    binSize = 0.01,
    ppm = 10,
    bw = 7)
```

We evaluate these settings, in particular the effect of `bw` on the first
example EIC.

```{r}
#| fig-cap: "Simulation of correspondence analysis on the first example EIC."
col_peak <- col_sample[chromPeaks(eic_1_adj)[, "sample"]]
plotChromPeakDensity(eic_1_adj, param = pdp, col = col_sample,
                     peakCol = col_peak, peakBg = paste0(col_peak, 40))
grid()
```

For that EIC the settings worked nicely. Simulating the correspondence for the
second example EIC.

```{r}
#| fig-cap: "Simulation of correspondence analysis on the second example EIC."
col_peak <- col_sample[chromPeaks(eic_2_adj)[, "sample"]]
plotChromPeakDensity(eic_2_adj, param = pdp, col = col_sample,
                     peakCol = col_peak, peakBg = paste0(col_peak, 40))
grid()
```

Also for the second EIC all chromatographic peaks got grouped into the same
feature. Ideally, correspondence parameters should also be simulated on more
complex signals, e.g. regions with co- or closely eluting compounds. We below
expand the retention time window for the *m/z* slice of the first example EIC.

```{r}
#| fig-cap: "Simulation of correspondence analysis on an expanded RT window."
a <- chromatogram(mse, mz = mzr_1, rt = c(30, 150))

col_peak <- col_sample[chromPeaks(a)[, "sample"]]
plotChromPeakDensity(a, param = pdp, col = col_sample,
                     peakCol = col_peak, peakBg = paste0(col_peak, 40))
grid()
```

There seem to be signal from 3 different compounds in that *m/z* slice. Using
the settings above, in particular with `bw = 7` the apparently different
chromatographic signals at about 105 and 115 seconds are grouped into the same
LC-MS feature as indicated with the grey rectangle in the plot above. Unless we
want that these signals are grouped together, we need to reduce the value of
`bw`. Below we simulate a correspondence using `bw = 3`.

```{r}
#| fig-cap: "Effect of changing to `bw = 3` for the correspondence analysis."
pdp@bw <- 3
plotChromPeakDensity(a, param = pdp, col = col_sample,
                     peakCol = col_peak, peakBg = paste0(col_peak, 40))
grid()
```

With `bw = 3` we successfully grouped the signals into 3 distinct features. We
next evaluate whether with these updated setting we would still group the signal
from the second example EIC into a single feature.

```{r}
#| fig-cap: "Effect of `bw = 3` on correspondence results of the second example EIC."
col_peak <- col_sample[chromPeaks(eic_2_adj)[, "sample"]]
plotChromPeakDensity(eic_2_adj, param = pdp, col = col_sample,
                     peakCol = col_peak, peakBg = paste0(col_peak, 40))
grid()
```

All chromatographic peaks for this region were grouped into a single region. We
thus proceed and use these settings for the correspondence analysis of the full
data set.

```{r}
#' perform correspondence analysis on the full data set
mse <- groupChromPeaks(mse, param = pdp)
```

It is advised to check the results again after correspondence analysis, also to
evaluate the impact of the other settings like `binSize` and `ppm` and assure
validity of the defined LC-MS features. Ideally, a larger number of
EICs/features should be checked. We below extract the first example EIC again
and evaluate the correspondence results for it. Here it is important to set
`simulate = FALSE` to show the actual results.

```{r}
#| fig-cap: "Correspondence results for the first example EIC."
eic_1 <- chromatogram(mse, rt = rtr_1, mz = mzr_1)

#' plot the actual correspondence results by setting `simulate = FALSE`
col_peak <- col_sample[chromPeaks(eic_1)[, "sample"]]
plotChromPeakDensity(eic_1, col = col_sample, peakCol = col_peak,
                     peakBg = paste0(col_peak, 40),
                     simulate = FALSE)
grid()
```

All chromatographic peaks were grouped into the same feature. The retention time
(`"rtmed"`) of the feature is indicated with a dashed vertical line. We next
evaluate the results also on the second example EIC.

```{r}
#| fig-cap: "Correspondence results for the second example EIC."
eic_2 <- chromatogram(mse, rt = rtr_2, mz = mzr_2)

#' plot the actual correspondence results by setting `simulate = FALSE`
col_peak <- col_sample[chromPeaks(eic_2)[, "sample"]]
plotChromPeakDensity(eic_2, col = col_sample, peakCol = col_peak,
                     peakBg = paste0(col_peak, 40),
                     simulate = FALSE)
grid()
```

Also for this retention time region, chromatographic peaks were grouped into a
single feature. At last we evaluate the expanded retention time region of the
*m/z* range of the first example EIC.

```{r}
#| fig-cap: "Correspondence results for an *m/z* slice with multiple closely eluting compounds."
a <- chromatogram(mse, mz = mzr_1, rt = c(30, 150))

col_peak <- col_sample[chromPeaks(a)[, "sample"]]
plotChromPeakDensity(a, col = col_sample, peakCol = col_peak,
                     peakBg = paste0(col_peak, 40),
                     simulate = FALSE)
grid()
```

For this region, chromatographic peaks were grouped into 3 distinct features.

The results from the correspondence analysis can be extracted from the result
object using the `featureDefinitions()` and `featureValues()` functions. The
former returns the *definition* of the LC-MS features, i.e., their *m/z* and
retention time, the latter the actual abundance estimates in the different
samples. We below count the number of features that were defined by the
correspondence analysis:

```{r}
featureDefinitions(mse) |>
    nrow()
```

A quite large number of features was defined. This is mostly due to the
setting for the `minFraction` parameter, which required a chromatographic peak
to be detected in only 2 out of the 3 replicates for each individual sample to
define, and combine them into, a feature.

We can now extract the abundance estimates of all features with the
`featureValues()` function. Below we extract this data matrix and assign the
unique sample name as column names (by default the MS data file name is used).

```{r}
fvals <- featureValues(mse, method = "sum")
colnames(fvals) <- sampleData(mse)$sample_desc
head(fvals)
```

Columns in this abundance matrix are samples, rows features. For the present
data set there are a large number of missing values. A missing value indicates
failure to detect a chromatographic peak for the *m/z* - retention time region
of a feature in a sample. This can have multiple reasons: the compound might
simply not be present in the sample or the original signal is too noisy, to low
in abundance, or does not fit the expected shape for the peak detection
algorithm to identify a chromatographic peak.

Below we calculate and plot the number of missing values per sample.

```{r}
#| fig-cap: "Number of missing feature values per sample."

#' determine the number of missing values per sample and plot them
nas <- apply(fvals, MARGIN = 2, function(z) sum(is.na(z)))

barplot(nas, main = "Number of missing values", col = col_sample)
```

The lowest number of missing values is present in the *PPL_R1* sample.

## Gap-filling

To reduce the number of missing values and avoid data imputation we perform
*gap-filling*: in samples with a missing value for a feature (i.e., in which no
chromatographic peak was detected) we integrate all intensities measured within
the *m/z* - retention time area of the feature. By default, this area is defined
as the 25% to 75% quantile of the lower, respectively upper *m/z* (and retention
time) boundary of all chromatographic peaks of a feature. As an example we
calculate this area with the `featureArea()` function for the first 6 features:

```{r}
featureArea(mse,
            mzmin = function(z) quantile(z, probs = 0.25, na.rm = TRUE),
            mzmax = function(z) quantile(z, probs = 0.75, na.rm = TRUE),
            rtmin = function(z) quantile(z, probs = 0.25, na.rm = TRUE),
            rtmax = function(z) quantile(z, probs = 0.75, na.rm = TRUE),
            features = rownames(featureDefinitions(mse))[1:4])
```

Thus, for gap-filling, missing values are replaced with the integrated signal
measured by the MS instrument within these *m/z* - retention time boundaries.

We below perform the gap-filling on the full data set.

```{r}
#' configure and perform gap-filling
cpap <- ChromPeakAreaParam(minMzWidthPpm = 10)
mse <- fillChromPeaks(mse, param = cpap, chunkSize = 4L)
```

We extract the feature values again and determine the number of missing values.

```{r}
fvals <- featureValues(mse, method = "sum")
colnames(fvals) <- sampleData(mse)$sample_desc
head(fvals)
```

```{r}
#| fig-cap: "Number of missing feature values per sample after gap-filling."

#' determine the number of missing values per sample and plot them
nas <- apply(fvals, MARGIN = 2, function(z) sum(is.na(z)))

barplot(nas, main = "Number of missing values", col = col_sample)
```

A considerable amount of values could thus be *rescued*.


# Extract fragment spectra for LC-MS features

After preprocessing, we next identify and extract the MS2 spectra for the
defined LC-MS features. We use the `featureSpectra()` method, that identifies
for each of the features' chromatographic peaks MS2 spectra (within the same
sample!) with their retention time and precursor *m/z* within the retention time
and *m/z* range of the chromatographic peak.

```{r}
#' identify MS2 spectra for features
ms2 <- featureSpectra(mse)
ms2
```

We can have multiple, or no, MS2 spectra per feature:

```{r}
#' count the number of MS2 spectra per feature
ms2_count <- table(ms2$feature_id)

#' the number of LC-MS features with at least one MS2 spectrum:
length(ms2_count)

#' the average number of MS2 spectra per feature:
mean(ms2_count)
```

We next inspect some of the identified MS2 spectra.

```{r}
#| fig-cap: "MS2 spectra for one LC-MS feature."
#| fig-width: 9.5
#| fig-height: 9.5
#' select MS2 spectra for the first feature
a <- ms2[ms2$feature_id == ms2$feature_id[1]]

#' plot the spectra
plotSpectra(a)
```

All MS2 spectra look similar - we next calculate also a pairwise similarity
between them and visualize the results as a heatmap.

```{r}
#| fig-cap: "Pairwise spectra similarity between MS2 spectra of one feature."
#| fig-width: 9.5
#| fig-height: 9.5
#' calculate dot product similarity
sim <- compareSpectra(a, ppm = 10, tolerance = 0)

pheatmap(sim)
```

Similarity between all MS2 spectra is very high (above 0.92).

We next combine the individual MS2 spectra of a feature to a single *consensus*
spectrum.

::: {.callout-note collapse="true"}
## Options to combine spectra

Multiple spectra can be combined into a single spectrum using the
`combineSpectra()` function which provides a large number of options and
parameters for this task, including the possibility to provide a custom function
to combine the fragment peaks. By default (with parameter `peaks = "union"`),
the combined spectrum contains **all** mass peaks of all input spectra. The
option `peaks = "intersect"` allows to retain only peaks that are present in a
certain proportion of input spectra. Parameters `ppm` and `tolerance` define the
required similarity in the peaks' *m/z* value to be considered *the same*. The
groups of spectra to be combined are specified with parameter `f` and a
potential splitting of the input `Spectra` object for parallel processing with
parameter `p`.
:::

As an example, we combine MS2 spectra keeping only mass peaks present in at
least 75% of input spectra.

```{r}
#' define consensus spectra per feature
ms2_cons <- combineSpectra(ms2, f = ms2$feature_id,
                           p = rep(1, length(ms2)),
                           peaks = "intersect",
                           ppm = 10, minProp = 0.75)
ms2_cons
```

We have thus now one consensus spectrum per feature. A summary of the numbers of
peaks per consensus spectrum is shown below.

```{r}
#' overview on the number of peaks per spectrum
quantile(lengths(ms2_cons))
```

The consensus spectrum for the first feature is shown below.

```{r}
#| fig-cap: "Consensus spectrum for the first feature."
plotSpectra(ms2_cons[1], lwd = 2)
grid()
```

We next filter the data and remove spectra with a single fragment peak.

```{r}
#' remove spectra with a single fragment peak
ms2_cons <- ms2_cons[lengths(ms2_cons) > 1]
ms2_cons
```

::: {.callout-note collapse="true"}
## Additional spectra processing options

The [*Spectra*](https://bioconductor.org/packages/Spectra) package would provide
many additional functions and options to process, scale or clean spectra. As an
alternative, through the
[*SpectriPy*](https://bioconductor.org/packages/SpectriPy) package, it would
also be possible to apply Python-based functionality from e.g. the *matchms*
Python library to `Spectra` objects.
:::

At last we visualize the select data (i.e. features with associated MS2 spectra)
in the *m/z* - retention time space. We use the `featureArea()` function to get
the feature boundaries and draw them as rectangles. The associated MS2 spectra
(their precursor *m/z* and retention time value) are added as individual data
points.

```{r}
#| fig-cap: "Feature areas (grey rectangles) and associated MS2 spectra (blue points) in the retention time - *m/z* space."
#| fig-width: 9.5
#| fig-height: 9.5
#' define the feature boundaries
fa <- featureArea(mse, features = ms2_cons$feature_id)

#' plot feature areas as rectangles
plot(NA, NA, xlim = range(fa[, c("rtmin", "rtmax")]),
     ylim = range(fa[, c("mzmin", "mzmax")]),
     xlab = "retention time", ylab = "m/z")
grid()
rect(xleft = fa[, "rtmin"], xright = fa[, "rtmax"],
     ybottom = fa[, "mzmin"], ytop = fa[, "mzmax"],
     border = "#00000080")
#' add precursor m/z and retention times of MS2
points(ms2_cons$rtime, ms2_cons$precursorMz,
       cex = 0.5, col = "#0000ff40")
```

Thus, with a little bit of R coding we can easily create customized data
visualizations.


# Formatting and exporting data for FBMN

We next export the feature abundance matrix and the fragment spectra for
feature-based molecular networking with GNPS. Similar to the original workflow
[@johannes_rainer_2025_17293665] described in [@nothias_feature-based_2020] we
write the feature matrix to a tabulator delimited text file and the associated
MS2 spectra to a file in Mascot Generic File (MGF) format. A future version of
the workflow might use the mzTab-M format for the data exchange.

We first compile the feature abundance matrix and export that to a txt file.

```{r}
#' get feature definitions
fdef <- featureDefinitions(mse)[, c("mzmed", "mzmin", "mzmax",
                                    "rtmed", "rtmin", "rtmax")]
#' combine with the feature value table
fvals <- cbind(Row.names = rownames(fdef), fdef, fvals)

#' restrict the feature abundance matrix to features with MS2 spectra
fvals <- fvals[ms2_cons$feature_id, ]

#' export the data
write.table(fvals, "xcms_ms2_features.txt", sep = "\t",
            quote = FALSE, row.names = FALSE)
```

We next reformat the information in the MS2 spectra restricting to data required
by GNPS. The respective functionality is at present provided in the
[xcms-gnps-tools](https://github.com/jorainer/xcms-gnps-tools) GitHub repo.

```{r}
#' load functions for GNPS-specific spectra formatting
source("https://raw.githubusercontent.com/jorainer/xcms-gnps-tools/master/customFunctions.R")
ms2_cons <- formatSpectraForGNPS(ms2_cons)
```

And finally we export the MS2 spectra in MGF format.

```{r}
#' export the MS2 spectra in MGF format
export(ms2_cons, backend = MsBackendMgf(),
       file = "xcms_ms2_spectra.mgf")
```

::: {.callout-note collapse="true"}
## Spectra data export formats

The MGF format is only loosely defined with many different *dialects* being
used. The [*MsBackendMgf*](https://bioconductor.org/packages/MsBackendMgf)
package supports renaming or specifying spectra variables (metadata) for export
to the MGF format. In addition, it would also allow to export also additional
peak information, such as chemical formulas for individual fragments. As an
alternative, also other export formats would be supported for `Spectra` objects,
provided by packages such as the
[*MsBackendMsp*](https://bioconductor.org/packages/MsBackendMsp) or
[*MsBackendMassbank*](https://bioconductor.org/packages/MsBackendMassbank). Upcoming
formats, such as *specLib*, *mzPeak* or the updated mzTab-M format will be
supported in future.

:::



# Summary

- R-based data analysis workflows allow data set specific, tailored, analysis of
  LC-MS data.
- The *xcms* R package for LC-MS data preprocessing is tightly integrated into a
  broader ecosystem of R packages.
- The quarto system would also allow combining R and Python functionality into
  the same workflow document with the
  [SpectriPy](https://github.com/RforMassSpectrometry/SpectriPy) R-package
  translating between R and Python MS data structures.

# Session information

The R version and package versions used:

```{r}
sessionInfo()
```

# References
